Storm Surges

### 2014-05-27 RoughWork/manuscript.tex
Outline of intro.

\section{Introduction}\label{sec:intro}
\citep{masson2004modelling} %testing a citation
%To be included: NEMO overview, tides, rivers, bathymetry, atmospheric forcing, vertical/lateral mixing, boundary conditions, grid.

\section{Model Configuration}\label{sec:config}
%A section about how we configured the model for SoG 
We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the biogeochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and biogeochemical coupling make it a useful tool for studying the complex physics and biogeochemical interactions within the Strait of Georgia. This work focuses on validating the physical set up of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include biogeochemical coupling and data assimilation.

@@

The curvilinear orthogonal numerical grid is divided into 400 by 900 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 400 m by 500 m. The 40 vertical $z$-levels are stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10m the grid is stretched gradually with a maximum grid spacing of 27 m at the lowest layer. At the bottom boundary, partial $z$-levels are utilized in order to limit large changes in bathymetry across grid cells \citep{madec2008nemo}.

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.png}
\caption{Model domain including bathymetry, rivers (*), and storm surge locations (o) of interest.}\label{fig:domain}
\end{figure}

\subsection{Boundary conditions and sub-grid scale processes}
%lateral and bottom friction, open boundaries, eta surface
The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents, temperature and salinity climatologies, and the sea surface height anomaly. The details of these forcing conditions are described below. At coastal boundaries, the partial slip boundary condition, an approximation of the no slip boundary condition, is used. The partial slip boundary condition allows one to include the frictional effects of lateral boundaries without the restrictive resolution required to represent the lateral boundary layer under no slip conditions. A lateral eddy viscosity of 20 m $^2$ s $^{-1}$ parametrizes horizontal friction and the lateral eddy diffusivity is 20.5 m $^2$ s $^{-1}$ . At the ocean surface, meteorological conditions and fresh water input from rivers are included and described in more detail below. Bottom friction is represented by a quadratic law for the bottom momentum flux with drag coefficient $C_D = 5\times 10^{-3}$. Vertical turbulence and mixing is calculated through the $k-\epsilon$ configuration of the generic length scale (GLS) turbulence closure \citep{umlauf2003generic} with background vertical eddy viscosity and diffusivity set to $1\times10^{-4}$ and $1\times10^{-5}$ m$^2$ s$^{-1}$ respectively. Details on the NEMO implementation of the partial slip lateral boundary condition, quadratic bottom friction law, and GLS turbulence closure scheme are provided by \citep{madec2008nemo}.

@@

\subsubsection{Tidal forcing} 
The model was forced by tidal elevations and currents at the Juan de Fuca and Johnstone Strait boundaries. Tidal heights and currents for eight tidal constituents (K1 ,O1, P1, Q1, M2, K2, N2, S2) at grid points along the Juan de Fuca boundary were extracted from Webtide, an online web prediction model for the northeast Pacific Ocean, which is based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M2 and K1 constituents. Additionally, O1 and S2 elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide.

@@

The model relaxed to the forced temperature and salinity over the 10 grid points (about 5km) closest to the open boundaries, using the NEMO flow relaxation scheme \citep{madec2008nemo}. %other references>
The tidal forcing and sea surface height was used in the barotropic velocity forcing which used the NEMO Flather scheme \citep{madec2008nemo}. %other references?
The baroclinic velocities at the boundary were set to be equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain.

@@

\subsection{Tidal evaluation}
The model was initially evaluated qualitatively by comparing patterns of tidal amplitude and phase to results from \citep{foreman1995tidal}. For example, the amphidromic dome around Victoria was produced in the $M_2$ results, as well as the monotonic decrease in $M_2$ amplitude moving northwards along the Strait of Georgia. Initially, the Johnstone Strait boundary was closed, but modelled $M_2$ amplitudes were too small compared to measured amplitudes... TBC

@@

>>>>

\section{Introduction}\label{sec:intro}

%some general intro paragraph here

%my attempt at physical oceanopgraphy SoG- Susan feel free to edit.
The Strait of Georgia is a strongly stratified, semi-enclosed body of water located between Vancouver Island and the mainland of British Columbia. It is part of a larger system of waterways collectively known as the Salish Sea and is connected to the Pacific Ocean via the Strait of Juan de Fuca and Johnstone Strait. Several features related to the physical oceanopgraphy of this region render it difficut to model numerically. The outlfow from the Fraser River in the late spring is the primary source fresh water within the Strait of Georgia, and leads to large density gradients in the summer months. Additionally, strong tidal mixing through the San Juan and Gulf Islands produces horizontal density gradients that separate the saline waters of the Strait of Juan de Fuca and the fresher water of the Strait of Georgia. It is important to accurately represent the vertical mixing in this region as it sets the rate of export of fresh water through estuarine circulation. The strong and fast tidal currents through the narrow channels of the north, such as Discovery Passage and Johnstone Strait, pose a particular challenge for numerical models. Yet, several modelling efforts for this region exist, ranging in choices of grid structure, mixing parametrizations, and domain size and extent.

%overview oF SoG/Salish Sea models

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damaged caused by storm surges, a natural hazard arising from the combination of a strong storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  Additionally, there is a small contribution to increased sea level due to the thermal expansion of water during warmer years set by the El Nino Southern Oscillation (ENSO) \citep{abeys2011extreme}. %more on past events, significance

%storm surge modelling

%overview of paper: model configuration, model evaluation, storm surge hindcasts.

\section{Model Configuration}\label{sec:config}
%To be included: NEMO overview, tides, rivers, bathymetry, atmospheric forcing, vertical/lateral mixing, boundary conditions, grid.

%A section about how we configured the model for SoG 
We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the biogeochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and biogeochemical coupling make it a useful tool for studying the complex physics and biogeochemical interactions within the Strait of Georgia. This work focuses on validating the physical set up of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include biogeochemical coupling and data assimilation.

@@

The curvilinear orthogonal numerical grid is divided into 398 by 898 by 40 grid cells, which results in an almost uniform horizontal resolution with grid spacing approximately 440 m by 500 m. The 40 vertical $z$-levels are stretched gradually in order to achieve higher resolution in the surface layer, with 1 m vertical grid spacing down to about 10 m in depth. Below 10m the grid is stretched gradually with a maximum grid spacing of 27 m at the lowest layer. At the bottom boundary, partial $z$-levels are utilized in order to limit large changes in bathymetry across grid cells \citep{madec2008nemo}.

\begin{figure}[h]
\centering
\includegraphics[scale=0.5]{Figures/bathy.png}
\caption{Model domain including bathymetry, rivers (*), and storm surge locations (o) of interest.}\label{fig:domain}
\end{figure}
%image needs some work: colour? labels for JdF, SoG, cities...

\subsection{Boundary conditions and sub-grid scale processes}
%lateral and bottom friction, open boundaries, eta surface
The model includes two open boundaries that connect to the Pacific Ocean, the western boundary of the Strait of Juan de Fuca as well as Johnstone Strait at the north, both of which are forced with eight tidal constituents, temperature and salinity climatologies, and the sea surface height anomaly. The details of these forcing conditions are described below. At coastal boundaries, the partial slip boundary condition, an approximation to no slip, is used. The partial slip boundary condition allows one to include the frictional effects of lateral boundaries without the restrictive resolution required to represent the lateral boundary layer under no slip conditions. A lateral eddy viscosity of 20 m $^2$ s $^{-1}$ parametrizes horizontal friction and the lateral eddy diffusivity is 20.5 m $^2$ s $^{-1}$ . At the ocean surface, meteorological conditions and fresh water input from rivers are included and described in more detail below. Bottom friction is represented by a quadratic law for the bottom momentum flux with drag coefficient $C_D = 5\times 10^{-3}$. Vertical turbulence and mixing is calculated through the $k-\epsilon$ configuration of the generic length scale (GLS) turbulence closure \citep{umlauf2003generic} with background vertical eddy viscosity and diffusivity set to $1\times10^{-4}$ and $1\times10^{-5}$ m$^2$ s$^{-1}$ respectively. Details on the NEMO implementation of the partial slip lateral boundary condition, quadratic bottom friction law, and GLS turbulence closure scheme are provided by \citet{madec2008nemo}.

@@

\subsubsection{Tidal forcing} 
The model was forced by eight tidal constituents (K1 ,O1, P1, Q1, M2, K2, N2, S2) at the Juan de Fuca and Johnstone Strait boundaries. Tidal heights and currents at grid points along the Juan de Fuca boundary were extracted from Webtide, an online web prediction model for the northeast Pacific Ocean, which is based on work by \citet{foreman2000webtide}. The Johnstone Strait boundary was forced with current and elevation tidal harmonics measured and calculated by \citet{thomson1980johnstone} for the major M2 and K1 constituents. Additionally, O1 and S2 elevation harmonics from their measurements were employed. The remaining constituents were extrapolated from Webtide.

@@

The model is relaxed to the forced temperature and salinity over the 10 grid points (about 5km) closest to the open boundaries, using a flow relaxation scheme \citep{engedahl1995use}. %other references?
The tidal forcing and sea surface height was used in the barotropic velocity forcing which used the NEMO Flather scheme \citep{flather1994storm, madec2008nemo}.
The baroclinic velocities at the boundary were set to be equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain.

@@

\subsection{Tidal evaluation}
The model was initially evaluated qualitatively by comparing patterns of tidal amplitude and phase to results from \citep{foreman1995tidal}. For example, the amphidromic dome around Victoria was produced in the $M_2$ results, as well as the monotonic increase in $K_1$ amplitude moving northwards along the Strait of Georgia. Initially, the Johnstone Strait boundary was closed, but modelled $M_2$ amplitudes were too small compared to measured amplitudes... TBC

@@

\subsection{Stratification}
